A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s

### 2018-11-06 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Add reference to Intel CSA about dark-silicon

In order to process the data, the 4 largest LHC experiments have written
more than 20 million lines of program code over the last 15 years
\cite{1742-6596-898-7-072013}. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

@@

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon~\cite{martin2014,IntelCSA}.

@@

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy to genomics,
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases
\cite{47224,Shanahan:2015:LSD:2783258.2789993,Armbrust:2015:SSR:2723372.2742797}.

@@

\begin{itemize}
\item
  Reviewing the physics models' assumptions, approximations, and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project~\cite{FCC}.
\item
  Redesigning, developing, and commissioning detector simulation
  toolkits to be more efficient when executed on current vector CPUs and
  emerging new architectures, including GPUs, where use of SIMD
  vectorisation is vital; this includes porting and optimising the
  experiments' simulation applications to allow exploitation of large
  HPC facilities.
\item
  Exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques. An area of investigation is common frameworks
  for fast tuning and validation.
\item
  Developing, improving and optimising geometry tools that can be shared
  am\-ong experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent.
\item
  Developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pileup).
\item
  Revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments.
\end{itemize}

@@

The reconstruction of raw detector data and simulated data, and its
processing in real time, represent a major component of today's
computing requirements in HEP. Advances in the capabilities of
facilities and future
experiments bring the potential for a dramatic increase in physics
reach, at the price of increased event complexities and rates. It is
therefore essential that event reconstruction algorithms and software
triggers continue to evolve so that they are able to efficiently exploit
future computing architectures, and deal with the increase in data rates
without loss of physics. Projections into future, e.g., at HL-LHC conditions,
show that without significant changes in approach or algorithms
the increase in resources needed would be incompatible with
the expected budget.

@@

At the same time, trigger systems for next-generation experiments are
evolving to be more capable, both in their ability to select a wider
range of events of interest for the physics programme, and their ability
to stream a larger rate of events for further processing. ATLAS and CMS
both target systems where the output of the hardware trigger system is
increased by an order of magnitude over the current capability, up to 1
MHz~\cite{CERN-LHCC-2015-020,CMSCollaboration:2015zni}. In LHCb~\cite{CERN-LHCC-2014-016}
and ALICE~\cite{Buncic:2011297}, the full collision
rate (between 30 to 40 MHz for typical LHC proton-proton operations)
will be streamed to real-time or near-real-time software trigger
systems. The increase in event complexity also brings a ``problem'' of
an overabundance of signals to the experiments, and specifically to the
software trigger algorithms. The evolution towards a genuine real-time
analysis of data has been driven by the need to analyse more signal than
can be written out for traditional processing, and technological
developments that enable this without reducing the analysis sensitivity
or introducing biases.

@@

Scientific questions are answered by analysing the data obtained from
suitably designed experiments and comparing measurements with
predictions from models and theories. Such comparisons are typically
performed long after data taking, but can sometimes also be executed in
near-real time on selected samples of reduced size.

@@

An important new development in the analysis domain has been the
emergence of new analysis tools coming from industry and open-source
projects (e.g. Jupyter notebooks~\cite{http://jupyter.org/}, the scikit-learn package~\cite{scikit-learn}), and this presents new opportunities for improving the HEP
analysis software ecosystem. The HEP community is very interested in
using these software tools, together with established components, in an
interchangeable way. The main challenge will be to enable new
open-source tools to be plugged in dynamically to the existing ecosystem
and to provide mechanisms that allow the existing and new components to
interact and exchange data efficiently. To improve our ability to
analyse much larger datasets, R\&D will be needed to investigate file
formats, compression algorithms, and new ways of storing and accessing
data for analysis and to adapt workflows to run on future computing
infrastructures.

Reproducibility is the cornerstone of scientific results. It is
currently difficult to repeat most HEP analyses in exactly the manner
they were originally performed. This difficulty mainly arises due to the
number of scientists involved, the large number of steps in a typical
HEP analysis workflow, and the complexity of the analyses themselves. A
challenge specific to data analysis and interpretation is tracking the
evolution of relationships between all the different components of an
analysis, i.e. the provenance of each step.

Reproducibility of scientific results goes in hand with the need to preserve both the data and the software.
Section~\ref{data-and-software-preservation}} develops this latter topic where the FAIR principles of data management are embraced.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark~\cite{Spark}, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as
TensorFlow~\cite{abadi2016tensorflow}, Dask~\cite{rocklin2015dask}, Pachyderm~\cite{thepachydermteam}, Blaze~\cite{wiebe2014blaze}, Parsl~\cite{babuji_yadu_2017_853492}, and Thrill~\cite{bingmann2016thrill}.
This approach can complement the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

@@

Another active area regarding analysis in the world outside HEP is the
switch to a functional or declarative programming model, as for example
provided by Scala~\cite{Scala} in the Spark environment. This allows scientists to
express the intended data transformation as a query on data. Instead of
having to define and control the ``how'', the analyst declares the
``what'' of their analysis, essentially removing the need to define the
event loop in an analysis, and leave it to underlying services and
systems to optimally iterate over events. It appears that these
high-level approaches will allow abstraction from the underlying
implementations, allowing the computing systems more freedom in
optimising the utilisation of diverse forms of computing resources. R\&D
is already under way, e.g., TDataFrame~\cite{TDataFrame} in ROOT, and
this needs to be continued with the ultimate goal of establishing a
prototype functional or declarative programming paradigm.

@@

\begin{itemize}
\item
  Enable new open-source software tools to be plugged in dynamically to
  the existing ecosystem, and provide mechanisms to dynamically exchange
  parts of the ecosystem with new components.
  \item
    Prototype a comprehensive set of mechanisms for interacting and
    exchanging data between new open-source tools and the existing
    analysis ecosystem.
\item
  Complete an advanced prototype of a low-latency response,
  high-capacity analysis facility, incorporating fast caching
  technologies to explore a query-based analysis approach and
  open-source cluster-management tools. It should in particular include
  an evaluation of additional storage layers, such as SSD storage and
  NVRAM-like storage, and cloud and Big Data orchestration systems.
\item
  Expand support of Python in our ecosystem with a strategy for ensuring
  long-term maintenance and sustainability. In particular in ROOT, the
  current Python bindings should evolve to reach the ease of use of native Python modules.
\item
  Develop a prototype based on a functional or declarative programming
  model for data analysis.
\item
  Conceptualise and prototype an analysis ``Interpretation Gateway'',
  including data repositories, e.g., HEPData~\cite{HEPData,HEPDataRepo}, and
  analysis preservation and reinterpretation tools.
\end{itemize}

@@

With the advent of more powerful hardware, particularly GPUs and ML dedicated
processors, as well as more performant ML
algorithms, the ML toolset will be used to develop application software
that could potentially, amongst other things:

@@

The R\&D roadmap presented here is based on the preliminary work done in
recent years, coordinated by the IML, which will remain the main
forum to coordinate work in ML in HEP and ensure the proper links
with the data science communities. The following programme of work is
foreseen.

@@

This increasingly dynamic environment for resources, particularly CPU,
must be coupled with a highly reliable system for data storage and a
suitable network infrastructure for delivering this data to where it
will be processed. While CPU and disk capacity is expected to increase
by respectively 15\% and 25\% per year for the same cost~\cite{CERN-COST-EST}, the trends of
research network capacity increases show a much steeper growth, such as
two orders of magnitude from now to HL-LHC times~\cite{Roberts:17}. Therefore, the
evolution of the computing models would need to be more network centric.

In the network domain, there are new technology developments, such as
Software Defined Networks (SDNs), which enable user-defined high
capacity network paths to be controlled via experiment software, and
which could help manage these data flows~\cite{Blikra:2221659}.
Some projects already started
to explore the potential of these technologies~\cite{OSiRIS} but a
considerable R\&D is required to prove their utility and practicality.
In addition,
the networks used by HEP are likely to see large increases in traffic
from other science domains.

@@

\begin{itemize}
\item
  Understand better the relationship between the performance and costs
  of the WLCG system, and how it delivers the necessary functionality to
  support LHC physics. This will be an ongoing process, started by the
  recently formed System Performance and Cost Modeling Working Group~\cite{SPCMWG},
  and aims to provide a quantitative assessment for any proposed
  changes.
\item
  Define the functionality needed to implement a federated data centre
  concept (``data lake'') that aims to reduce the operational cost of
  storage for HL-LHC, and at the same time better manage network
  capacity, whilst maintaining the overall CPU efficiency. This would
  include the necessary qualities of service, and options for regionally
  distributed implementations, including the ability to flexibly respond
  to model changes in the balance between disk and tape. This work
  should be done in conjunction with the existing Data Organisation,
  Management and Access Working Group~\cite{WLCG-DOMA} to evaluate the impact of the
  different access patterns and data organisations envisaged.
\item
  Building upon the experience of projects currently exploring SDN
  potential, define the role for this technology in managing data transfers and
  access and the integration strategy into experiment frameworks.
\item
  Establish an agreement on the common data management functionality
  that is required by experiments, targeting a consolidation and a lower
  maintenance burden. The intimate relationship between the management
  of elements in storage systems and metadata must be recognised. This
  work requires coordination with the Data Processing Frameworks
  Working Group. It
  needs to address at least the following use cases:

  \begin{itemize}
  \item
    processing sites that may have some small disk cache, but do not
    manage primary data;
  \item
    fine grained processing strategies that may enable processing of
    small chunks of data, with appropriate bookkeeping support;
  \item
    integration of heterogeneous processing resources, such as HPCs and
    clou\-ds.
  \end{itemize}
\item
  Explore scalable and uniform means of workload scheduling, which
  incorporate dynamic heterogenous resources, and the capabilities of
  finer grained processing that increases overall efficiency. The
  optimal scheduling of special workloads that require particular
  resources is clearly required.
\item
  Contribute to the prototyping and evaluation of a quasi-interactive
  analysis facility that would offer a different model for physics
  analysis, but would also need to be integrated into the data and
  workload management of the experiments. This is work to be done in
  collaboration with groups working on new data analysis models.
\end{itemize}

@@

The working group will also work towards a more convenient access to geometry and
event data, through a client-server interface~\cite{1742-6596-898-7-072015}. In collaboration with the
Data Access and Management Working Group, an API or a service to deliver streamed
event data would be designed.

@@

The workshop that was held as part of the CWP process
(\emph{HSF Visualization Workshop}, see Appendix \ref{appendix-a---list-of-workshops})
was felt to be
extremely useful for exchanging knowledge between developers in
different experiments, fostering collaboration and in bringing in ideas
from outside the community. This will now be held as an annual event and
will facilitate work on the common R\&D plan.

@@

To maximise productivity, it is very beneficial to use development tools
that are not HEP-specific. There are many open source projects that are
of similar scale to large experiment software stacks and standard tools
are usually well documented. For source control HEP has generally chosen
to move to \emph{git}~\cite{git},
which is very welcome, as it also brings an
alignment with many open source projects and commercial organisations.
A major benefit that has come with this technical choice is the use
of social coding sites, such as \emph{GitHub} ~\cite{GitHub}
and \emph{GitLab}~\cite{GitLab}, where
code sharing and code review are far superior compared to previous
solutions.
Likewise, \emph{CMake}~\cite{CMake}
is widely used for the builds of software
packages, both within HEP and outside. Packaging many build products
together into a software stack is an area that still requires close
attention with respect to active developments (the HSF has an active
working group here).

@@

Ongoing activities and short-term projects, to complete by 2020,
include the following:

@@

Longer term projects, to conclude by 2022, include the following:

@@

The above-mentioned steps are consistent with the
FAIR data principles that are increasingly being mandated by funding
agencies~\cite{FAIRdata}.

@@

The evolution in the security domain requires the HEP community to work in
collaboration with the various national security organisations and
policy groups, building on many relationships that are already
established.

@@

The existing technologies leveraged by identity federations, e.g., the
Security Assertion Markup Language (SAML), have not supported non-web
applications historically. There is momentum within the wider community
to develop next-generation identity federations~\cite{OICFederation}
that natively support a
wider range of clients. In the meantime there are several viable interim
solutions that are able to provision users with the token required to
access a service (such as X.509) transparently, translated from their
home organisation identity.

Although non-X509 federated identity provides a potential solution for our
challenges in Authentication, Authorisation should continue to be
tightly controlled by the HEP community. Enabling Virtual Organisation
(VO) membership for federated credentials and integrating such a
workflow with existing identity vetting processes is a major topic
currently being worked on, in particular within the WLCG community.
Commercial clouds and HPC centres have fundamentally different access
control models and technologies from our grid environment. We shall need
to enhance our access control model to ensure compatibility and
translate our grid-based identity attributes into those consumable by
such services.

@@

Over the next decade, it is expected that considerable changes will be
made to address security in the domains highlighted above. The
individual groups, in particular those mentioned above, working in the
areas of trust and policies, operational security, authentication and
authorisation, and technology evolutions, are driving the R\&D
activities. These groups are generally much broader than just the HEP
community. The list below summarises the most important actions:

@@

\begin{itemize}
\item
By 2020:
\begin{itemize}
\item
  Ensure that ongoing efforts in trust frameworks are sufficient to
  raise the level of confidence in non-X509 federated identities to the
  equivalent of X.509, at which stage they could be a viable alternative
  to both grid certificates and CERN accounts.
\item
  Participate in setting directions for the future of identity
  federations, through the FIM4R~\cite{FIM4R} community.
\end{itemize}
\item
By 2022:
\begin{itemize}
\item
  Overhaul the current Authentication and Authorisation infrastructure,
  including Token Translation, integration with Community IdP-SP
  Proxies, and Membership Management tools. Enhancements in this area
  are needed to support a wider range of user identities for WLCG
  services.
\end{itemize}
\end{itemize}

@@

We know that in any future scenario development effort will be
constrained, so it is vital that successful R\&D projects provide
sustainable software for the future. It is important to emphasise that the goal
is to support the HEP physics programme in a cost effective manner, so the
deployment consequences of a particular technology choice or direction must be
understood with partners in distributed computing. In many areas it is recognised
that different experiments could have adopted common solutions,
reducing overall development effort and increasing robustness and
functionality. That model of duplicated development is not sustainable. We must
endeavour to achieve better coherence within HEP for future developments
to build advanced, open-source projects that can be shared and supported
in common. The HSF has already established itself as a forum that can
facilitate this. Establishing links outside of HEP, to other academic
disciplines, to industry, and to the computer science community, can
strengthen both the research and production phases of new solutions. We
should ensure that the best products are chosen, from inside and outside
HEP, and that they receive support from all parties, aiming at technical
excellence and economy of scale.

>>>>

In order to process the data, the 4 largest LHC experiments have written
more than 20 million lines of program code over the last 15 years. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

@@

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon~\cite{IntelCSA}.

@@

Particle physics is no longer alone in facing these massive data
challenges. Experiments in other fields, from astronomy to genomics,
will produce huge amounts of data in the future, and will need to
overcome the same challenges that we face, i.e., massive data handling and
efficient scientific programming. Establishing links with these fields
has already started. Additionally, interest from the computing science
community in solving these data challenges exists, and mutually
beneficial relationships would be possible where there are genuine
research problems that are of academic interest to that community and
provide practical solutions to ours. The efficient processing of massive
data volumes is also a challenge faced by industry, in particular the
internet economy, which developed novel and major new technologies under
the banner of \emph{Big Data} that may be applicable to our use cases.

@@

\begin{itemize}
\item
  Reviewing the physics models' assumptions, approximations, and
  limitations in order to achieve higher precision, and to extend the
  validity of models up to energies of the order of 100 TeV foreseen
  with the Future Circular Collider (FCC) project~\cite{FCC}.
\item
  Redesigning, developing, and commissioning detector simulation
  toolkits to be more efficient when executed on current vector CPUs and
  emerging new architectures, including GPUs, where use of SIMD
  vectorisation is vital; this includes porting and optimising the
  experiments' simulation applications to allow exploitation of large
  HPC facilities.
\item
  Exploring different Fast Simulation options, where the full detector
  simulation is replaced, in whole or in part, by computationally
  efficient techniques. An area of investigation is common frameworks
  for fast tuning and validation.
\item
  Developing, improving and optimising geometry tools that can be shared
  am\-ong experiments to make the modeling of complex detectors
  computationally more efficient, modular, and transparent.
\item
  Developing techniques for background modeling, including contributions
  of multiple hard interactions overlapping the event of interest in
  collider experiments (pileup).
\item
  Revisiting digitisation algorithms to improve performance and
  exploring opportunities for code sharing among experiments.
\item
  Recruiting, training, retaining human resources in all areas of
  expertise pertaining to the simulation domain, including software and
  physics.
\end{itemize}

@@

The reconstruction of raw detector data and simulated data, and its
processing in real time, represent a major component of today's
computing requirements in HEP. Advances in the capabilities of 
facilities and future
experiments bring the potential for a dramatic increase in physics
reach, at the price of increased event complexities and rates. It is
therefore essential that event reconstruction algorithms and software
triggers continue to evolve so that they are able to efficiently exploit
future computing architectures, and deal with the increase in data rates
without loss of physics. Projections into future, e.g., at HL-LHC conditions,
show that without significant changes in approach or algorithms
the increase in resources needed would be incompatible with the 
the expected budget.

@@

At the same time, trigger systems for next-generation experiments are
evolving to be more capable, both in their ability to select a wider
range of events of interest for the physics programme, and their ability
to stream a larger rate of events for further processing. ATLAS and CMS
both target systems where the output of the hardware trigger system is
increased by an order of magnitude over the current capability, up to 1
MHz~\cite{CERN-LHCC-2015-020,CMSCollaboration:2015zni}. In LHCb~\cite{CERN-LHCC-2014-016}
and ALICE~\cite{Buncic:2011297}, the full collision
rate (between 30 to 40 MHz for typical LHC proton-proton operations)
will be streamed to real-time or quasi-real-time software trigger
systems. The increase in event complexity also brings a ``problem'' of
an overabundance of signals to the experiments, and specifically to the
software trigger algorithms. The evolution towards a genuine real-time
analysis of data has been driven by the need to analyse more signal than
can be written out for traditional processing, and technological
developments that enable this without reducing the analysis sensitivity
or introducing biases.

@@

Scientific questions are answered by analysing the data obtained from
suitably designed experiments and comparing measurements with
predictions from models and theories. Such comparisons are typically
performed long after data taking, but can sometimes also be executed in
quasi-real time on selected samples of reduced size.

@@

An important new development in the analysis domain has been the
emergence of new analysis tools coming from industry and open source
projects, and this presents new opportunities for improving the HEP
analysis software ecosystem. The HEP community is very interested in
using these software tools, together with established components, in an
interchangeable way. The main challenge will be to enable new
open-source tools to be plugged in dynamically to the existing ecosystem
and to provide mechanisms that allow the existing and new components to
interact and exchange data efficiently. To improve our ability to
analyse much larger datasets, R\&D will be needed to investigate file
formats, compression algorithms, and new ways of storing and accessing
data for analysis and to adapt workflows to run on future computing
infrastructures.

Reproducibility is the cornerstone of scientific results. It is
currently difficult to repeat most HEP analyses in exactly the manner
they were originally performed. This difficulty mainly arises due to the
number of scientists involved, the large number of steps in a typical
HEP analysis workflow, and the complexity of the analyses themselves. A
challenge specific to data analysis and interpretation is tracking the
evolution of relationships between all the different components of an
analysis.

@@

One new model of data analysis, developed outside of HEP, maintains the
concept of sequential reduction, but mixes interactivity with
batch processing. These exploit new cluster management systems, most
notably Apache Spark, which uses open-source tools contributed both by
industry and the data-science community. Other products implementing the
same analysis concepts and workflows are emerging, such as TensorFlow,
Dask, Pachyderm, Blaze, Parsl, and Thrill. This approach can complement
the present and widely adopted Grid processing of datasets. It may
potentially simplify the access to data and the expression of
parallelism, thereby improving the exploitation of cluster resources.

@@

Another active area regarding analysis in the world outside HEP is the
switch to a functional or declarative programming model, as for example
provided by Scala in the Spark environment. This allows scientists to
express the intended data transformation as a query on data. Instead of
having to define and control the ``how'', the analyst declares the
``what'' of their analysis, essentially removing the need to define the
event loop in an analysis, and leave it to underlying services and
systems to optimally iterate over events. It appears that these
high-level approaches will allow abstraction from the underlying
implementations, allowing the computing systems more freedom in
optimising the utilisation of diverse forms of computing resources. R\&D
is already under way, e.g., TDataFrame~\cite{TDataFrame} in ROOT, and
this needs to be continued with the ultimate goal of establishing a
prototype functional or declarative programming paradigm.

@@

\begin{itemize}
\item
  Enable new open-source software tools to be plugged in dynamically to
  the existing ecosystem, and provide mechanisms to dynamically exchange
  parts of the ecosystem with new components.
\item
  Complete an advanced prototype of a low-latency response,
  high-capacity analysis facility, incorporating fast caching
  technologies to explore a query-based analysis approach and
  open-source cluster-management tools. It should in particular include
  an evaluation of additional storage layers, such as SSD storage and
  NVRAM-like storage, and cloud and Big Data orchestration systems.
\item
  Expand support of Python in our ecosystem with a strategy for ensuring
  long-term maintenance and sustainability. In particular in ROOT, the
  current Python bindings should evolve to reach the ease of use of native Python modules.
\item
  Prototype a comprehensive set of mechanisms for interacting and
  exchanging data between new open-source tools and the existing
  analysis ecosystem.
\item
  Develop a prototype based on a functional or declarative programming
  model for data analysis.
\item
  Conceptualise and prototype an analysis ``Interpretation Gateway'',
  including data repositories, e.g., HEPData~\cite{HEPData,HEPDataRepo}, and
  analysis preservation and reinterpretation tools.
\end{itemize}

@@

With the advent of more powerful hardware and more performant ML
algorithms, the ML toolset will be used to develop application software
that could potentially, amongst other things:

@@

The R\&D roadmap presented here is based on the preliminary work done in
recent years, coordinated by the HSF IML, which will remain the main
forum to coordinate work in ML in HEP and ensure the proper links
with the data science communities. The following programme of work is
foreseen.

@@

This increasingly dynamic environment for resources, particularly CPU,
must be coupled with a highly reliable system for data storage and a
suitable network infrastructure for delivering this data to where it
will be processed. While CPU and disk capacity is expected to increase
by respectively 15\% and 25\% per year for the same cost~\cite{CERN-COST-EST}, the trends of
research network capacity increases show a much steeper growth, such as
two orders of magnitude from now to HL-LHC times. Therefore, the
evolution of the computing models would need to be more network centric.

In the network domain, there are new technology developments, such as
Software Defined Networks (SDNs), which enable user-defined high
capacity network paths to be controlled via experiment software, and
which could help manage these data flows. These new technologies require
considerable R\&D to prove their utility and practicality. In addition,
the networks used by HEP are likely to see large increases in traffic
from other science domains.

@@

\begin{itemize}
\item
  Understand better the relationship between the performance and costs
  of the WLCG system, and how it delivers the necessary functionality to
  support LHC physics. This will be an ongoing process, started by the
  recently formed System Performance and Cost Modeling Working Group,
  and aims to provide a quantitative assessment for any proposed
  changes.
\item
  Define the functionality needed to implement a federated data centre
  concept (``data lake'') that aims to reduce the operational cost of
  storage for HL-LHC, and at the same time better manage network
  capacity, whilst maintaining the overall CPU efficiency. This would
  include the necessary qualities of service, and options for regionally
  distributed implementations, including the ability to flexibly respond
  to model changes in the balance between disk and tape. This work
  should be done in conjunction with the existing Data Organisation,
  Management and Access Working Group to evaluate the impact of the
  different access patterns and data organisations envisaged.
\item
  Establish an agreement on the common data management functionality
  that is required by experiments, targeting a consolidation and a lower
  maintenance burden. The intimate relationship between the management
  of elements in storage systems and metadata must be recognised. This
  work requires coordination with the Data Processing Frameworks
  Working Group. It
  needs to address at least the following use cases:

  \begin{itemize}
  \item
    processing sites that may have some small disk cache, but do not
    manage primary data;
  \item
    fine grained processing strategies that may enable processing of
    small chunks of data, with appropriate bookkeeping support;
  \item
    integration of heterogeneous processing resources, such as HPCs and
    clou\-ds.
  \end{itemize}
\item
  Explore scalable and uniform means of workload scheduling, which
  incorporate dynamic heterogenous resources, and the capabilities of
  finer grained processing that increases overall efficiency. The
  optimal scheduling of special workloads that require particular
  resources is clearly required.
\item
  Contribute to the prototyping and evaluation of a quasi-interactive
  analysis facility that would offer a different model for physics
  analysis, but would also need to be integrated into the data and
  workload management of the experiments. This is work to be done in
  collaboration with the Data Analysis and Interpretation Working Group.
\end{itemize}

@@

The working group will also work towards a more convenient access to geometry and
event data, through a client-server interface. In collaboration with the
Data Access and Management Working Group, an API or a service to deliver streamed
event data would be designed.

@@

The workshop that was held as part of the CWP process was felt to be
extremely useful for exchanging knowledge between developers in
different experiments, fostering collaboration and in bringing in ideas
from outside the community. This will now be held as an annual event and
will facilitate work on the common R\&D plan.

@@

To maximise productivity, it is very beneficial to use development tools
that are not HEP-specific. There are many open source projects that are
of similar scale to large experiment software stacks and standard tools
are usually well documented. For source control HEP has generally chosen
to move to \emph{git}~\cite{git},
which is very welcome, as it also brings an
alignment with many open source projects and commercial organisations.
Likewise, \emph{CMake}~\cite{CMake} 
is widely used for the builds of software
packages, both within HEP and outside. Packaging many build products
together into a software stack is an area that still requires close
attention with respect to active developments (the HSF has an active
working group here).

@@

Ongoing activities and short-term projects, include the following:

@@

Medium term projects include the following:

@@

The above-mentioned steps can be considered to be consistent with the
FAIR data principles that are increasingly being mandated by funding
agencies~\cite{FAIRdata}.

@@

@@

The existing technologies leveraged by identity federations, e.g., the
Security Assertion Markup Language (SAML), have not supported non-web
applications historically. There is momentum within the wider community
to develop next-generation identity federations that natively support a
wider range of clients. In the meantime there are several viable interim
solutions that are able to provision users with the token required to
access a service (such as X.509) transparently, translated from their
home organisation identity.

Although federated identity provides a potential solution for our
challenges in Authentication, Authorisation should continue to be
tightly controlled by the HEP community. Enabling Virtual Organisation
(VO) membership for federated credentials and integrating such a
workflow with existing identity vetting processes is a major topic
currently being worked on, in particular within the WLCG community.
Commercial clouds and HPC centres have fundamentally different access
control models and technologies from our grid environment. We shall need
to enhance our access control model to ensure compatibility and
translate our grid-based identity attributes into those consumable by
such services.

@@

Over the next decade, it is expected that considerable changes will be
made to address security in the domains highlighted above. The
individual groups, in particular those mentioned above, working in the
areas of trust and policies, operational security, authentication and
authorisation, and technology evolutions, are driving the R\&D
activities. The list below summarises the most important actions:

@@

\begin{itemize}
\item
By 2020:
\begin{itemize}
\item
  Ensure that ongoing efforts in trust frameworks are sufficient to
  raise the level of confidence in federated identities to the
  equivalent of X.509, at which stage they could be a viable alternative
  to both grid certificates and CERN accounts.
\item
  Participate in setting directions for the future of identity
  federations, through the FIM4R~\cite{FIM4R} community.
\end{itemize}
\item
By 2022:
\begin{itemize}
\item
  Overhaul the current Authentication and Authorisation infrastructure,
  including Token Translation, integration with Community IdP-SP
  Proxies, and Membership Management tools. Enhancements in this area
  are needed to support a wider range of user identities for WLCG
  services.
\end{itemize}
\end{itemize}

@@

We know that in any future scenario development effort will be
constrained, so it is vital that successful R\&D projects provide
sustainable software for the future. In many areas it is recognised 
that different experiments could have adopted common solutions, 
reducing overall development effort and increasing robustness and 
functionality. That model of duplicated development is not sustainable. We must
endeavour to achieve better coherence within HEP for future developments
to build advanced, open-source projects that can be shared and supported
in common. The HSF has already established itself as a forum that can
facilitate this. Establishing links outside of HEP, to other academic
disciplines, to industry, and to the computer science community, can
strengthen both the research and production phases of new solutions. We
should ensure that the best products are chosen, from inside and outside
HEP, and that they receive support from all parties, aiming at technical
excellence and economy of scale.
